Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2018-01-03 paper.tex
Add Francois' edits through Sec. III

\subsection{Trajectories}
\label{ssec:trajectories}
Each trajectory is uniquely labeled by a pair of vectors
giving some event on the trajectory, $x^\alpha$,
and the momentum at that event, $p_\beta$.
A family of trajectories shares the same $x^\alpha$; for example,
the family of emission trajectories is ($x^\alpha_e,p_\beta$),
and the family of observation trajectories is ($x^\alpha_o,p_\beta$).

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
\footnote{TODO: code these up}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \ell'^\gamma,
\end{align}
defining the specific energy density,
specific momentum density, and
specific radiation pressure tensor, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\footnote{TODO: relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos a)\,db$.
\footnote{TODO: consistent with Eqn.~\ref{eqn:def_direction}?}
We also make use of the number density and flux defined
\begin{align}
  \label{eqn:G}
  G(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta), \\
  \label{eqn:Ka}
  K^\mu(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu.
\end{align}

@@

We integrate these equations through the simulated spacetime
over which the following volume data are known:
the spacetime metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
the fluid velocity $u_i$,
Lorentz factor $W$,
density $\varrho$,
temperature $T$,
and electron fraction $Y_e$,
all defined in App.~\ref{sec:definitions}.
These fields are stored in spectral representation.
If they are computed via hydrodynamical simulation, they may be saved to disk
at specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a quasi-stationary solution to the
general relativistic hydrodynamics equations,
a single time is stored to disk, and no time interpolation is needed.
In this paper for simplicity and to limit computational memory loads,
we limit ourselves to quasistationary configurations
and use single time slices in every test case.

@@

Since no analytic form is known for the background field $\Phi$
interior to the star, we examine only the \emph{noscat} case,
with $\chi_s=0$ and $\chi=\chi_a^{*}$.
This scenario has been widely used in the literature as a test for
radiation codes
\cite{smit1997-two_moment,abdi2012-monte_carlo,fouc2015-m1_nsbh}.
We sample the distribution function over a uniform grid of
30 points in angle $\cos A \in [0.734,1]$ and
40 points in energy $\tilde{\varepsilon} \in [0,100]$~MeV.
Because of the step function in fluid variables at radius $R$,
we throttle the maximum step size at 1.25~km.

@@

The $\langle\tilde{\varepsilon}_{\bar{\nu}_e}\rangle^{\rm sc}$ prediction
in the \emph{gray} \emph{scat} treatment is approximately 30\%
larger than the the fiducial Monte Carlo estimate. This is
due to the large negative local chemical potential
$\eta_{\bar{\nu}_e}\sim-10$ in the scattering envelope.
As described in Sec.~\ref{ssec:sources_se},
we use $\eta_\nu$ in the \emph{gray} treatment to construct our
synthetic spectrum from total neutrino densities $J$ and $G$.
This error and our successful analysis using pinched spectra above,
points the way to future improvements to the \emph{gray} method:
making better assumptions about the spectra which don't depend upon
local fluid quantities.

\subsection{Hypermassive Neutron Star Remnant:
  Testing Everything}
\label{ssec:test_disk_comparison}
The merger of two neutron stars by gravitational wave emission produces a
postmerger configuration composed of a single neutron star surrounded by a disk.
Because of strong differential rotation and shock heating,
the remnant temporarily avoids
collapse to a black hole, even if its mass exceeds the threshold of dynamical
instability for a rigidly rotating neutron star \cite{duez2009-review}.
These objects, called hypermassive neutron stars,
may avoid collapse for $X\textrm{--}Y$~s,
\footnote{TODO: cite $X$, $Y$}
depending upon a number of physical factors including thermal pressure,
magnetic fields, and the microphysics of the fluid.

Such a configuration was modeled in \cite{fouc2016-m1_nsns} by
evolving fluid and spacetime through the
final inspiral and merger of two identical neutron stars of
isolated gravitational mass 1.2~$M_{\odot}$. The postmerger configuration
was simulated using a gray M1 transport scheme for the neutrinos, evolving the
energy density, number density, and energy flux in addition to the standard
fluid and metric variables for $\sim10\,{\rm ms}$ following merger
\cite{fouc2016-m1_evolve_n}.
The fluid was modeled using the LS220 equation of state \cite{latt1991-nuc_eos}.

@@

Each observer samples the distribution function over a uniform grid
in energy, cosine of polar angle, and azimuthal angle with extents
$\{N_{\tilde{\varepsilon}},N_A, N_B\}$ spanning the ranges
$\tilde{\varepsilon}\in(0,100)$~MeV,
$\cos A\in\left((\cos A)_{\rm min}, 1\right)$, and
$B\in[0,2\pi)$.
The integrals for fluxes of number and energy
(Eqns.~\ref{eqn:Ha}, \ref{eqn:Ka}, and \ref{eqn:J_H_S_eps_integrated})
become sums over all rays.
Taking these fluxes in the radial direction,
and using the midpoint rule to convert the integral to a sum, we have
\begin{align}
  \label{eqn:Kr_minkowski}
  \tilde{K}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^2 \cos A_m f_m,\\
  \label{eqn:Hr_minkowski}
  \tilde{H}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^3 \cos A_m f_m,
\end{align}
with $f_m \equiv f(\tilde{\varepsilon}_m,\cos A_m, B_m)$,
$\Delta \equiv \Delta\tilde{\varepsilon}\, \Delta(\cos A)\, \Delta B$,
$N \equiv N_{\tilde{\varepsilon}}\, N_A\, N_B$,
and $m$ the index labeling each ray.
Average energies are given by
$\langle\tilde{\varepsilon}\rangle(\theta)=H^r(\theta)/K^r(\theta)$.
We choose $N_{\tilde{\varepsilon}}=30$, $N_A=150$, and $N_B=20$.
To maintain high angular resolution, we only sample rays that pass within
approximately 120~km of the star's center by setting $(\cos A)_{\rm min}=0.88$.

@@

Tab.~\ref{tab:nsns_rt_vs_m1} compares the all-sky
luminosities and average energies from ray tracing and from
the M1 transport simulation,
which serves for qualitative comparison.
Even if M1 and ray tracing methods both provide faithful measurements
of all-sky luminosities, we do not expect perfect agreement since the two
treatments differ fundamentally. In addition to differences in the physics
such as the treatment of inelastic scattering,
M1 fluxes are integrated over the outer boundary of the finite-difference grid
at radii as small as 75~km,
whereas the ray tracing fluxes are integrated over a sphere at radius 250~km,
introducing a time lag between the two measurements of under a millisecond.
Furthermore uncertainties from the M1 simulation are around
15\% for energy luminosities and
10\% for average energies \cite[Sec.~A.6]{fouc2016-m1_evolve_n}.

>>>>

\subsection{Trajectories}
\label{ssec:trajectories}
Each trajectory is uniquely labeled by a pair of vectors
giving an event on the trajectory, $x^\alpha$,
and the momentum at that event, $p_\beta$.
To designate a family of intersecting trajectories, we keep constant either
the emission event $x^\alpha_e$ or
the observation event $x^\alpha_o$.

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \ell'^\gamma,
\end{align}
defining the specific energy density,
specific momentum density, and
specific radiation pressure tensor, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\footnote{TODO: relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos a)\,db$.
\footnote{TODO: consistent with Eqn.~\ref{eqn:def_direction}?}
We also make use of the specific number density and specific number flux defined
\begin{align}
  \label{eqn:G}
  G(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta), \\
  \label{eqn:Ka}
  K^\mu(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu.
\end{align}

@@

We integrate these equations through the simulated spacetime
over which the following volume data are known:
the spacetime metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
the fluid velocity $u_i$,
Lorentz factor $W$,
density $\varrho$,
temperature $T$,
and electron fraction $Y_e$,
all defined in App.~\ref{sec:definitions}.
These fields are computed in a preprocessing step before ray tracing
and stored in spectral representation.
If they are computed from a hydrodynamical simulation, they may be saved to disk
at either one or several specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a stationary solution to the
general relativistic hydrodynamics equations no time interpolation is needed.
In this paper for simplicity and to limit computational memory loads,
we use only stationary analytical solutions or
quasi-stationary configurations evolved in simulation,
thus using one time slice and no time interpolation in every case.

@@

Since no analytic form is known for the background field $\Phi$
interior to the star, we examine only the \emph{noscat} case,
with $\chi_s=0$ and $\chi=\chi_a^{*}$.
This scenario has been widely used in the literature as a test for
radiation codes
\cite{smit1997-two_moment,abdi2012-monte_carlo,fouc2015-m1_nsbh}.
We sample the distribution function over a uniform grid of
30 points in angle $\cos A \in [0.734,1]$ and
40 points in energy $\tilde{\varepsilon} \in [0,100]$~MeV.
Because of the discontinuity in fluid variables at radius $R$,
we limit the time step size to a maximum of $t_{\rm max}=1.25$~km,
since as the ray approaches the discontinuity in the homogeneous environment
outside the star, the adaptive time-stepper would otherwise
increase the step size far beyond the relevant fluid scales.

@@

The $\langle\tilde{\varepsilon}_{\bar{\nu}_e}\rangle^{\rm sc}$ prediction
in the \emph{gray} \emph{scat} treatment is approximately 30\%
larger than the the fiducial Monte Carlo estimate. This is
due to the large negative local chemical potential
$\eta_{\bar{\nu}_e}\sim-10$ in the scattering envelope.
As described in Sec.~\ref{ssec:sources_se},
we use $\eta_\nu$ in the \emph{gray} treatment to construct our
synthetic spectrum from total neutrino densities $J$ and $G$.
This error and our successful analysis using pinched spectra above,
points the way to future improvements to the \emph{gray} method:
making better assumptions about the spectra which are less sensitive
to local fluid quantities.

\subsection{Hypermassive Neutron Star Remnant:
  Testing Everything}
\label{ssec:test_disk_comparison}
The merger of two neutron stars by gravitational wave emission produces a
postmerger configuration composed of a single neutron star surrounded by a disk.
Because of strong differential rotation and shock heating,
the remnant may temporarily avoid collapse to a black hole,
even if its mass exceeds the threshold of dynamical
instability for a rigidly rotating neutron star \cite{duez2009-review}.
These objects, called hypermassive neutron stars,
may avoid collapse for thousands of seconds
\footnote{TODO: cite Rezzolla et Kumar 2014, Ravi et Lasky 2014},
depending upon a number of physical factors including thermal pressure,
magnetic fields, and the microphysics of the fluid.

Such a configuration was modeled in \cite{fouc2016-m1_nsns} by
evolving fluid and spacetime through the
final inspiral and merger of two identical neutron stars of
isolated gravitational mass 1.2~$M_{\odot}$. The configuration
was simulated using a gray M1 transport scheme for the neutrinos, evolving the
energy density, number density, and energy flux in addition to the standard
fluid and metric variables to $\sim11\,{\rm ms}$ following merger
\cite{fouc2016-m1_evolve_n}.
The fluid was modeled using the LS220 equation of state \cite{latt1991-nuc_eos}.

@@

Each observer samples the distribution function over a uniform grid
in energy, cosine of polar angle, and azimuthal angle with extents
$\{N_{\tilde{\varepsilon}},N_A, N_B\}$ spanning the ranges
$\tilde{\varepsilon}\in(0,100)$~MeV,
$\cos A\in\left((\cos A)_{\rm min}, 1\right)$, and
$B\in[0,2\pi)$.
The integrals for fluxes of number and energy
(Eqns.~\ref{eqn:Ha}, \ref{eqn:Ka}, and \ref{eqn:J_H_S_eps_integrated})
become sums over all rays.
Taking these fluxes in the radial direction,
and using the midpoint rule to convert the integral to a sum, we have
\begin{align}
  \label{eqn:Kr_minkowski}
  \tilde{K}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^2 \cos A_m f_m,\\
  \label{eqn:Hr_minkowski}
  \tilde{H}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^3 \cos A_m f_m,
\end{align}
with $f_m \equiv f(\tilde{\varepsilon}_m,\cos A_m, B_m)$,
$\Delta \equiv \Delta\tilde{\varepsilon}\, \Delta(\cos A)\, \Delta B$,
$N \equiv N_{\tilde{\varepsilon}}\, N_A\, N_B$,
and $m$ the index labeling each ray.
Average energies in the coordinate frame are given by
$\langle\tilde{\varepsilon}\rangle(\theta)=\tilde{H}^r(\theta)/\tilde{K}^r(\theta)$.
We choose $N_{\tilde{\varepsilon}}=30$, $N_A=150$, and $N_B=20$.
To maintain high angular resolution, we only sample rays that pass within
approximately 120~km of the star's center by setting $(\cos A)_{\rm min}=0.88$.

@@

Tab.~\ref{tab:nsns_rt_vs_m1} compares the all-sky
luminosities and average energies from ray tracing and from
the M1 transport simulation,
which serves for qualitative comparison.
Even if M1 and ray tracing methods both provide faithful measurements
of all-sky luminosities, we do not expect perfect agreement since the two
treatments differ fundamentally. In addition to differences in the physics
such as the treatment of inelastic scattering,
M1 fluxes are integrated over the outer boundary of the finite-difference grid
at radii ranging from 75~km to 200~km,
whereas the ray tracing fluxes are integrated over a sphere at radius 250~km,
introducing a time lag between some of the fluxes used in the measurements
of order one millisecond.
Additionally, uncertainties for the M1 simulation may be estimated from
comparisons between M1 methods to be around
15\% for energy luminosities and
10\% for average energies \cite[Sec.~A.6]{fouc2016-m1_evolve_n}.
